---
title: "About Me"
---
My name is Hong Zhuang. This is my first semester in the QMM program at Baruch College. I graduated with a Bachelor degree in Business Administration from Baruch College and a Master's degree in Science from CUNY, City College.\
Currently, I'm working as a staff accountant at a retail start-up. I am committed to become a risk analyst at a financial institution. I have passed CFA Level1, Level2 and FRM Level1. If you share the same career path, feel free to connect with me on [LinkedIn](https://www.linkedin.com/in/hongzhuang10/).


```{r}
#| echo: false
#| message: false
#| warning: false

if(!require("leaflet")){
    options(repos=c(CRAN="https://cloud.r-project.org"))
    install.packages("leaflet")
    stopifnot(require("leaflet"))
}

baruch_longitude <- -73.98333
baruch_latitude <- +40.75028

leaflet() |>
  addTiles() |>
  setView(baruch_longitude, baruch_latitude, zoom=17) |>
  addPopups(baruch_longitude, baruch_latitude, 
            "I am a Master's student at <b>Baruch College</b>!")
```

```{r}
#| include: false
1+1
```
--------------
Last Updated: `r format(Sys.time(), "%A %m %d, %Y at %H:%M%p")`