---
title: "Soft Rock Playlist 🎸"
output:
  html: default
  pdf: default
code-fold: true
toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

## Ultimate Playlist Nomination

The **Soft Rock Playlist** opens with mellow, acoustic-rich songs like *The Scientist* 
and *Photograph*, showing high acousticness and moderate energy. As we progress, 
there's a noticeable rise in energy and valence—peaking with track like *Toxic*.
However, the return to softer emotional tones in *Yellow* reflects a ‘rise and fall’ 
pattern rather than a linear build-up—echoing the emotional arc of a live concert set.

<iframe src="ultimate_playlist.html" width="100%" height="500px" style="border: none;"></iframe>

## Design Principles 

### Audio Features
Related song were identified based on key musical characteristics shared by the 
anchor songs **The Scientist** and **Viva La Vida** by *Coldplay*, such as matching 
key and similar ranges of **tempo**, **acousticness** and **danceability**.

### By Genre Identity
Songs by the rock band: *U2* were selected due to their similar
genre and sonic identity. Like *Coldplay*, *U2* is known for its atmospheric sound
and emotional resonance, making them a natural stylistic match.

### Spotlight on Hidden Gems
In addition to well-known tracks with a popularity score of 79 or above, a few tracks
with lower popularity score were included to add diversity and depth to the playlist.

## Data Acquisition: `songs` dataset

```{r}
# Load Relevant Packages
library(tidyverse)
library(dbplyr)
library(stringr)
library(ggplot2)
library(DT)
library(scales)
library(gt)
library(rvest)
library(purrr)
library(httr)
library(jsonlite)
library(knitr)
library(kableExtra)
library(ggthemes)
library(tidyr)
library(stringr)

# Task 1: function to load songs 
load_songs <- function(url){
  data_dir <- file.path("data", "mp03")
  file_name <- file.path(data_dir, "spotify_data.csv")
  
  if(!dir.exists(data_dir)){
    dir.create(data_dir, showWarnings=FALSE, recursive=TRUE)
  }
  
  if(!file.exists(file_name)){
    download.file(url, destfile = file_name, method = "curl")
  }
  
  # Read .csv file into R 
  songs <- read_csv(file_name)
  songs <- as.data.frame(songs)
  
  return(songs)
}

# Split the `artists` across multiple rows
clean_artist_string <- function(x){
    str_replace_all(x, "\\['", "") |> 
        str_replace_all("'\\]", "") |>
        str_replace_all("'", "")
}

# Call the function `load_songs` to acquire Spotify `songs` dataset
url <- "https://raw.githubusercontent.com/gabminamedez/spotify-data/refs/heads/master/data.csv"
songs <- load_songs(url)

# Clean up the `artists` column 
songs <- songs |> 
  separate_longer_delim(artists, ",") |>
  mutate(artist = clean_artist_string(artists)) |>
  select(-artists)

# Display the first 10 rows of `songs` data
datatable(head(songs, 10), options = 
            list(scrollX = TRUE, scrollY = 300, paging = TRUE))
```


## Data Acquisition: `playlist` dataset

```{r}
# Load relevant packages
library(httr)
library(tibble)

# Task 2: Write function to load Spotify million playlists 
load_playlists <- function(base_url){
  data_dir <- file.path("data", "mp03")
  
  if(!dir.exists(data_dir)){
    dir.create(data_dir, showWarnings=FALSE, recursive=TRUE)
  }
  
  start_indices <- seq(0, 999000, by = 1000)
  
  # Create an empty list to store data frames of playlists 
  playlists <- list()
  
  for (start in start_indices){
    end <- start + 999
    json_file_name <- sprintf("mpd.slice.%d-%d.json", start, end)
    file_url <- paste0(base_url, json_file_name)
    file_name <- file.path(data_dir, json_file_name)
    
    # First check if the JSON file already exists to handle duplicate download
    if(!file.exists(file_name)){
      response <- HEAD(file_url)
      if(status_code(response) == 200){ 
        download.file(file_url, destfile = file_name, method = "curl", quiet = TRUE) # Download only if the raw file exists on Github
      }else{
        next
      }
    }
    
    if(file.exists(file_name)){
      json_data <- fromJSON(file_name)
      if(! is.null(json_data$playlists)){            # It handles the missing playlists from JSON file or null playlists
        playlists_df <- as_tibble(json_data$playlists)
        playlists[[length(playlists) + 1]] <- playlists_df
      }
    }
  }
  
  final_playlists <- bind_rows(playlists)
  return(final_playlists)
}


# Call the function `load_playlists` to acquire Spotify million playlists
base_url <- "https://raw.githubusercontent.com/DevinOgrady/spotify_million_playlist_dataset/main/data1/"
playlists <- load_playlists(base_url)

glimpse(playlists)
```

## Data Cleaning and Formatting

```{r}
# Select the desired columns
playlists <- playlists |>
  select(name, pid, num_followers, tracks)

# Task 3: Expands the list-column "tracks" for each row
playlists <- playlists |>
  unnest(tracks)

# Rename columns for readability and reorder the columns 
playlists <- playlists |>
  rename(
    playlist_name = name,
    playlist_id = pid,
    playlist_followers = num_followers,
    playlist_position = pos,
    artist_id = artist_uri,
    track_id = track_uri,
    album_id = album_uri,
    duration = duration_ms
  ) |>
  select(
    playlist_name,
    playlist_id,
    playlist_position,
    playlist_followers,
    artist_name,
    artist_id,
    track_name,
    track_id,
    album_name,
    album_id,
    duration
  )

# Clean up the ID columns
strip_spotify_prefix <- function(x){
  library(stringr)
  str_extract(x, ".*:.*:(.*)", group=1)
}

playlists <- playlists |>
  mutate(across(c(artist_id, track_id, album_id), strip_spotify_prefix))

# Display the first 10 rows of `playlists` data
datatable(head(playlists, 10), options = 
            list(scrollX = TRUE, scrollY = 300, paging = TRUE))
```


## Initial Exploration

```{r, include=FALSE}
create_gt_table <- function(x){
  # Rename columns 
  colnames(x) <- colnames(x) |>
    str_replace_all("_", " ") |>
    str_to_title()
  
  # Bold these columns
  bold_cols <- c("Playlist", "Track Name", "Name", "Playlist Name")
  cols_to_bold <-intersect(names(x), bold_cols)
  
  # Identify text and numeric columns for alignment and number formatting below 
  text_cols <- names(x)[map_lgl(x, ~ is.character(.x) || is.factor(.x))]
  numeric_cols <- names(x)[map_lgl(x, is.numeric)]
  
  gt_table <- x |>
    gt() |>
    tab_options(
      table.width = "100%"
    ) |>
    tab_style(
      style = list(
        cell_text(weight = "bold")
      ),
      locations = cells_column_labels(everything())
    ) |>
    tab_style(
      style = cell_text(weight = "bold", color = "#264653"),
      locations = cells_body(columns = all_of(cols_to_bold))
    ) |>
    cols_align(
      align = "left",
      columns = all_of(text_cols)
    ) |>
    cols_align(
      align = "center",
      columns = all_of(numeric_cols)
    )
  
  return(gt_table)
}
```


There are **1,200,590** distinct tracks and **173,604** distinct artists in 
Spotify playlist dataset.

```{r}
# number of distinct tracks
playlists |>
  select(track_id) |>
  n_distinct() 

# number of distinct artists 
playlists |>
  select(artist_id) |>
  n_distinct()
```

### The Top 5 Most Popular Tracks

```{r}
# Count the frequency of each track's appearance across playlists
top_5 <- playlists |>
  group_by(track_id) |>
  summarize(
    track_name = first(track_name),
    artist_name = first(artist_name),
    n_appearances = n()
    ) |>
  slice_max(n_appearances, n = 5)

create_gt_table(top_5)
```

### Missing Tracks

Among the top five most popular tracks, **One Dance** and **Broccoli (feat.Lil Yachty)**
are the two tracks missing from the `songs` dataset. 

```{r}
missing_tracks <- anti_join(top_5, songs, by = c("track_id" = "id")) |> 
  select(- n_appearances)
  
create_gt_table(missing_tracks)
```

### Most Danceable Track

**Funky Cold Medina** is the most "danceable" track, with a danceability 
score of 0.988, and it appears in 211 playlists. 

```{r}
# Find the most danceable track 
most_danceable <- songs |>
  select(name, id, danceability) |>
  slice_max(danceability, n = 1)

# Count the frequency of appearance in the playlists dataset
most_danceable_playlist_count <- playlists |>
  inner_join(most_danceable, by = c("track_id" = "id")) |>
  count(track_id) |>
  rename(playlist_count = n)

most_danceable <- most_danceable |>
  inner_join(most_danceable_playlist_count, by = c("id" = "track_id"))

create_gt_table(most_danceable)
```


### Playlist with the Longest Average Track Length

The playlist **Mixes** has the longest average track length, with an average duration 
of 64 minutes.

```{r}
longest_playlist <- playlists |>
  group_by(playlist_id, playlist_name) |>
  summarize(
    avg_length = round(mean(duration, na.rm = TRUE)/60000, 2)
  ) |>
  ungroup() |>
  slice_max(avg_length, n = 1)

create_gt_table(longest_playlist)
```

### The Most Popular Playlist

The most popular playlist on Spotify is **Breaking Bad** with 53,519 followers.

```{r}
most_popular_playlist <- playlists |>
  select(playlist_name, playlist_id, playlist_followers) |>
  distinct() |>
  slice_max(playlist_followers, n = 1)

create_gt_table(most_popular_playlist)
```


## Visually Identify Characteristics of Popular Songs

### Define Popularity Benchmark

```{r}
# Inner join the two datasets: `playsists` and `songs_unique` 
playlist_song_features <- playlists |>
  inner_join(songs, by = c("track_id" = "id", "artist_name" = "artist"))

# I chose the song "Viva La Vida" by Coldplay.  
# It has a popularity score of 79.

popular_songs <- playlist_song_features |>
  filter(popularity >= 79) 
```


### Do Popular Songs Appear in More Playlists?

From the scatter plot below, it seems there is no strong linear correlation 
between the popularity of a song and the number of its playlist appearances.
This is further supported by the correlation coefficient of 0.42, which indicates
a weak linear relationship.

```{r, message=FALSE}
song_popularity <- playlist_song_features |>
  select(track_id, playlist_id, popularity) |>
  group_by(track_id) |>
  summarize(playlist_freq = n(),
            avg_popularity = mean(popularity, na.rm = TRUE)
            )

ggplot(song_popularity, aes(x = playlist_freq, y = avg_popularity)) +
  geom_point(color = "#3A86FF") +
  labs(title = "Track Popularity vs. Playlist Count",
       x = "Number of Playlist Appearances",
       y = "Popularity Value"
       ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 18, hjust = 0.5)
  )
```

```{r}
cor(song_popularity$playlist_freq, song_popularity$avg_popularity)
```

### The Number of Most Popular Songs by Year

The bar chart below displays the years when the most popular songs were released.
The earliest year in which songs with a popularity value above 79 were released 
in 1967.

The number of popular songs doesn't vary much between year 1967 and year 2013. 
However, we see a spike in year 2017, when the number of popular songs reached 
approximately 150,000. 

```{r}
popular_songs |>
  count(year) |>
  ggplot(aes(x = factor(year), y = n/1000)) +
  geom_col(fill = "#3A86FF") +
  scale_fill_tableau(palette = "Tableau 20") +
  labs(
    title = "Number of Top Songs by Year",
    x = "Release Year",
    y = "Number of Popular Songs (in thousands)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", size = 18, hjust = 0.5)
  )
```

### Average Danceability of Most Popular Songs by Year

The "dancebility" peaked in year 1982, with a value of 0.92.

```{r}
library(RColorBrewer)

popular_songs |>
  group_by(year) |>
  summarize(avg_danceability = round(mean(danceability, na.rm = TRUE), 2)) |>
  ggplot(aes(x = factor(year), y = avg_danceability, fill = factor(year))) +
  geom_col(color = "grey", show.legend = FALSE) +  # Adding grey borders for clarity
  geom_text(aes(label = avg_danceability), vjust = -0.3, size = 2.5) +  # Adding value labels above bars
  scale_fill_manual(
    values = rep(brewer.pal(12, "Set3")[5], n_distinct(popular_songs$year)) 
  ) +
  labs(
    title = "Danceability Value of Most Popular Songs by Year",
    x = "Year",
    y = "Average Danceability"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),  # Adjusted year text for readability
    plot.title = element_text(face = "bold", size = 18, hjust = 0.5)  # Center title
  )

```

### The Most Represented Decade on User Playlist

There appears to be a clear trend showing an increase in playlist count by decade.
The 2010s saw the highest number of playlists created -more than double 
the count from the 2000s. 

```{r}
library(glue)
bar_chart <- function(value, color = "#3a86ff", height = "30px") {
  glue("<span style='display: inline-block; height: {height}; width: {value}%; background-color: {color}; border-radius: 4px;'></span>")
}

playlist_count_by_decade <- popular_songs |>
  mutate(
    decade = (year %/% 10) * 10
  ) |>
  group_by(decade) |>
  summarize(playlist_count = n_distinct(playlist_id))

total_count <- sum(playlist_count_by_decade$playlist_count)

playlist_count <- playlist_count_by_decade$playlist_count

playlist_count_by_decade_with_bar <- playlist_count_by_decade |>
  mutate(
    percent = round(playlist_count/total_count * 100, 2),
    bar = map(percent, ~gt::html(bar_chart(.x)))
  ) |>
  select(-percent)

gt(playlist_count_by_decade_with_bar) |>
  cols_label(
    decade = "Decade",
    playlist_count = "Number of Playlists Created",
    bar = "Percent"
  ) |>
  fmt_number(
    columns = playlist_count,
    sep_mark = ",",
    decimals = 0
  ) |>
  cols_align(
    align = "center",
    columns = c("decade", "playlist_count")
  ) |>
  cols_align(
    align = "left",
    columns = bar
  ) |>
  tab_options(
    table.width = "100%"
  ) |>
  tab_style(
    style = cell_text(color = "black", weight = "bold"),
    locations = list(
      cells_column_labels(everything())
    )
  )
```

### Polar Plot of Key Frequency

As shown in the plot,  the majority of songs are in key 1 and mode 0.

```{r}
key_frequency <- popular_songs |>
  select(track_id, mode, key) |>
  group_by(mode, key) |>
  summarise(key_freq = n())

ggplot(key_frequency, aes(x=factor(key), y = key_freq, fill = factor(mode)))+
  geom_bar(stat = "identity", position = "stack", width = 1) +
  coord_polar(start = 0) +
  scale_fill_manual(
    values = c("0" = "#43a2ca", "1" = "#a8ddb5"),
    name = "Mode"
  ) +
  labs(
    title = "Song Frequency by Musical Key and Mode",
    x = "Musical Key (0 = C, 1 = C♯/D♭, ..., 11 = B)",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    panel.grid = element_blank(),
    axis.text.x = element_text(size = 10),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
```

### The Most Popular Track Length

The majority of the popular songs have track lengths in the range of 3 and 4.5 minutes.

```{r}
popular_songs |>
  mutate(duration_in_min = round(duration / 60000, 2)) |>
  ggplot(aes(x = duration_in_min)) +
  geom_histogram(binwidth = 0.5, fill = "#3a86ff", color = "white") +
  labs(
    title = "Distribution of Popular Track Length (in Minutes)",
    x = "Duration (Minutes)",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5)
  )
```

### Track Lengths of the Top 20 Most Frequently Included Popular Songs

```{r, message=FALSE}
# Find the most included tracks by counting the number of distinct playlists 
the_most_included_songs <- popular_songs |>
  group_by(track_id, track_name) |>
  summarize(
    n_playlist = n_distinct(playlist_id),
    avg_length = round(mean(duration, na.rm = TRUE)/60000, 2) # Convert the duration from ms to minutes
  ) |>
  ungroup() |>
  slice_max(n_playlist, n = 20)

the_most_included_songs |>
ggplot(aes(x = reorder(track_name, -n_playlist), y = avg_length, fill = track_name)) +
  geom_col() +
  scale_fill_tableau(palette = "Tableau 20") +
  labs(
    title = "Track Length of the Top 20 Most Frequently Included Tracks",
    x = "Track Name",
    y = "Track Length (in minutes)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold")) 
```

### Coldplay's Popular Songs

There are four songs by Coldplay with a popularity score of 79 or higher. *The Scientist*
is included in the most playlists, followed by *Viva La Vida*, *Fix You* and *Yellow*.
Again, we can see there is no direct relationship between the number of playlist 
appearance and the popularity value.

```{r}
# Count the number of playlists in which each of Coldplay's popular song is included
coldplay <- popular_songs|>
  filter(artist_name == "Coldplay") |>
  group_by(track_name) |>
  summarize(
    n_playlists = n_distinct(playlist_id),
    popularity = mean(popularity, na.rm = TRUE))

# Add a new column to display bars based on popularity values
coldplay <-coldplay |>
  mutate(
    bar = map(popularity, ~gt::html(bar_chart(.x)))
  ) |>
  select(-popularity)

# Visualize the data using gt
gt(coldplay) |>
  cols_label(
    track_name = "Track Name",
    n_playlists = "Number of Playlists ",
    bar = "Popularity"
  ) |>
  fmt_number(
    columns = n_playlists,
    sep_mark = ",",
    decimals = 0
  ) |>
  cols_align(
    align = "center",
    columns = c("track_name", "n_playlists")
  ) |>
  cols_align(
    align = "left",
    columns = bar
  ) |>
  tab_options(
    table.width = "90%"
  ) |>
  tab_style(
    style = cell_text(color = "black", weight = "bold"),
    locations = list(
      cells_column_labels(everything())
    )
  )
```


### The Distribution of `popularity` for Popular Songs by Decade

The distribution of `popularity` is almost even by decade, oscillating in the 
range of 79 to 82. Only in 2010's are there two songs with values above 90. 

```{r}
library(RColorBrewer)

popular_songs |>
  mutate(
    decade = (year %/% 10) * 10
  ) |>
  group_by(decade) |>
  ggplot(aes(x = factor(decade), y = popularity, fill = factor(decade))) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Set3") +
  labs(
    title = "Popularity Distribution by Decade",
    x = "Decade",
    y = "Popularity"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 18, hjust = 0.5)
    ) 
```

## Building a Playlist from Anchor Songs

```{r, echo=FALSE}
# Select desired columns
data_formatting <- function(x) {
  x <- x |>
    select(artist_name, track_name, popularity, duration,
           year, acousticness, danceability, energy, instrumentalness, liveness,
           loudness, speechiness, tempo, valence, mode, key, explicit
           ) |>
    arrange(desc(popularity))
  
  return(x)
}

create_datatable <- function(x){
  datatable(x, options = 
            list(scrollX = TRUE, scrollY = 300, paging = TRUE))
}
```

### Anchor Songs and Their Features

My two anchor songs are: **The Scientist** and **Viva La vida** by *Coldplay*.
The table below shows their features, which will be used to find candidate songs 
to include in the ultimate playlist.

```{r}
# Chose anchor songs 
anchor_songs <- playlist_song_features |>
  filter(track_name %in% c("The Scientist", "Viva La Vida") & popularity >= 79) |>
  distinct(track_name, .keep_all = TRUE)

anchor_songs <- data_formatting(anchor_songs)
create_datatable(anchor_songs)
```

### Most Common Songs in Playlists with Anchor Songs

To find the most common songs that appear in playlists containing the anchor songs,
I first identify the distinct playlists where anchor songs appear.
Then, I count how often each song appears in those playlists and select the top 
20 most common ones.

```{r}
# First, find the list of distinct playlists where anchor songs appear
anchor_song_playlist <- playlist_song_features |>
  filter(track_id %in% c("75JFxkI2RXiU7L9VXzMkle", "1mea3bSkSGXuIRvnydlB5b")) |> 
  select(playlist_name, playlist_id) |>
  distinct()

# Find the songs appear in the playlists above by inner_join the two datasets by "playlist_id"
top20_common_songs <- playlist_song_features |>
  inner_join(anchor_song_playlist, by = "playlist_id") |> 
  group_by(track_id) |>
  summarize(
    track_name = first(track_name),
    freq = n_distinct(playlist_id)     # Count the number of playlist appearances for each track_id
  ) |>
  ungroup() |>
  filter(! track_name %in% c("The Scientist", "Viva La Vida") ) |>
  slice_max(freq, n = 20)

# Select the top 20 common songs
top20_common_songs <- playlist_song_features |>
  inner_join(top20_common_songs, by = "track_id") |>
  rename(
    track_name = track_name.x
  ) |>
  select(-track_name.y, -freq) |>
  distinct(track_id, .keep_all = TRUE)

top20_common_songs <- data_formatting(top20_common_songs)
create_datatable(top20_common_songs)
```

### Tracks with Matching Key and in Similar Tempo Range

There are 348 songs that share the same key as the anchor songs, with tempo ranging 
from 138 to 146 BPM. I filtered the top 10 based on the popularity score.

```{r}
top10_songs_in_same_key <- playlist_song_features|>
  filter(key == 5, tempo >= 138 & tempo <= 146, 
         ! track_name %in% c("The Scientist", "Viva La Vida")) |>
  distinct(track_name, .keep_all = TRUE) |>
  arrange(desc(popularity)) |>
  slice_head(n = 10)

top10_songs_in_same_key <- data_formatting(top10_songs_in_same_key)
create_datatable(top10_songs_in_same_key)
```


### Other Songs by Coldplay

```{r}
other_songs_by_coldplay <- playlist_song_features|>
  filter(artist_name == "Coldplay",! name %in% c("The Scientist", "Viva La Vida")) |>
  distinct(track_id, .keep_all = TRUE)

other_songs_by_coldplay <- data_formatting(other_songs_by_coldplay)
create_datatable(other_songs_by_coldplay)
```

### Songs Released in the Same Years as Anchors

There are 11 songs released in the same years as the two anchor songs, filtered 
based on two specific musical characteristics: acousticness and danceability.

```{r}
# The anchor songs were released in 2002 and 2008 respectively
songs_released_same_year <- playlist_song_features |>
  filter(year %in% c(2002, 2008),
         acousticness >= 0.7 & acousticness <= 0.75,
         danceability >= 0.46 & danceability <= 0.58,
         ! track_name %in% c("The Scientist", "Viva La Vida")
         ) |>
  distinct(track_id, .keep_all = TRUE)

songs_released_same_year <- data_formatting(songs_released_same_year)
create_datatable(songs_released_same_year)
```

### Songs by Another Artist with a Similar Genre

```{r}
songs_by_U2 <- playlist_song_features |>
  filter(artist_name == "U2") |>
  distinct(track_name, .keep_all = TRUE)

songs_by_U2 <- data_formatting(songs_by_U2)
create_datatable(songs_by_U2)
```

### Candidates for the Ultimate Playlist

```{r}
candidate_songs <- bind_rows(top20_common_songs, top10_songs_in_same_key, 
                             other_songs_by_coldplay, songs_released_same_year,
                             anchor_songs)
create_datatable(candidate_songs)
```


## Create the Ultimate Playlist

Using the principles outlined at the beginning, the ultimate playlist 
- **Soft Rock Playlist** was produced.

- **Acousticness**: *Dream* and *The Scientist* lean acoustic, adding softness 
  and emotional intimacy.
- **Energy**: Broad range from mellow (*Dream*, 0.248) to high-energy like *Toxic* 
  (0.838). This contrast supports dynamic flow within the playlist-rising and falling
  emotional and sonic intensity.
- **Tempo**: The playlist moves from slower to more upbeat tracks, creating a 
  range of moods.
- **Valence**: The playlist rises and falls emotionally, starting withe reflective,
  low-valence songs and transitionning toward brighter, more energetic tracks.

```{r}
# the least popular songs from `top20_common_songs`
ultimate_playlist <- top20_common_songs |> tail(2)

# Select the first two songs from the rest of the candidate song data frames below
list_of_candidate_song_dataframes <- list(anchor_songs, 
                                          top10_songs_in_same_key, 
                                          other_songs_by_coldplay, 
                                          songs_released_same_year, 
                                          songs_by_U2
                                          )

to_add <-  list_of_candidate_song_dataframes |>
  lapply(head, 2) |>
  bind_rows()

# Final ultimate playlist
ultimate_playlist <- ultimate_playlist |>
  bind_rows(to_add)

# Reorder by the columns by `acousticness`
ultimate_playlist <- ultimate_playlist |>
  arrange(desc(acousticness))
```

```{r}
ggplot(ultimate_playlist, aes(x = fct_reorder(track_name, acousticness, .desc = TRUE),  
                              y = acousticness, group = 1)) +
  geom_line(color = "#3a86ff", size = 1) +  
  geom_point(color = "#4895ef", size = 2) + 
  labs(
    title = "Acoustic Levels in the Soft Rock Playlist",
    x = "Track Name (Ordered by Acousticness)",
    y = "Acousticness Value"
  ) +
  theme_minimal() +  # Minimal theme for cleaner visuals
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),  # Rotate x-axis text for readability
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5)  # Center and bold the title
  )

```



```{r}
ggplot(ultimate_playlist, aes(x = fct_reorder(track_name, acousticness, .desc = TRUE),  
                              y = energy, group = 1)) +
  geom_line(color = "#ffd60a", size = 1) +
  geom_point(color = "#ffb703", size = 2) +  
  labs(
    title = "Energy Levels in the Soft Rock Playlist",
    x = "Track Name (Ordered by Acousticness)",  
    y = "Energy Value"
  ) +
  theme_minimal() +  # Minimal theme for cleaner visuals
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),  # Rotate x-axis text for readability
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5)  # Center and bold the title
  )
```


```{r}
ggplot(ultimate_playlist, aes(x = fct_reorder(track_name, acousticness, .desc = TRUE),  
                              y = valence, group = 1)) +
  geom_line(color = "#8338ec", size = 1) +  
  geom_point(color = "#7209b7", size = 2) +  
  labs(
    title = " Valence Levels in the Soft Rock Playlist",
    x = "Track Name (Ordered by Acousticness)", 
    y = "Valence Value"
  ) +
  theme_minimal() +  # Minimal theme for cleaner visuals
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),  # Rotate x-axis text for readability
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5)  # Center and bold the title
  )
```


```{r}
ggplot(ultimate_playlist, aes(x = fct_reorder(track_name, acousticness, .desc = TRUE),  
                              y = tempo, group = 1)) +
  geom_line(color = "#ff0054", size = 1) +  
  geom_point(color = "#e5383b", size = 2) +  
  labs(
    title = " Tempo Levels in the Soft Rock Playlist",
    x = "Track Name (Ordered by Acousticness)", 
    y = "Tempo Value"
  ) +
  theme_minimal() +  # Minimal theme for cleaner visuals
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),  # Rotate x-axis text for readability
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5)  # Center and bold the title
  )
```




